Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
edit discussion 3

We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs). While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress germline transcripts[@scandagliaLossKdm5cCauses2017]. Although KDM5C is enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of EpiLC germline-enriched DEGs. One notable DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells that promotes meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Retinoic acid can only activate _Stra8_ expression when DAZL is present, which is a direct target of KDM5C and is aberrantly transcribed and translated in _Kdm5c_-KO EpiLCs. This indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C regulation through activation by other ectopic germline programs. Consistent with this idea, many _Kdm5c_-KO EpiLC germline DEGs are important for early germ cell development and meiosis while those expressed in the mature _Kdm5c_-KO brain are involved in late sperm development. Altogether, this indicates ectopic germline programs are, to some extent, progressing through germ cell developmental stages over the course of _Kdm5c_-KO development.

- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
- Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654

@@

\clearpage
# Notes
## Things to do
- Move dazl to new figure if other staining works
- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
- Add cell culture concentrations to methods

@@

- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
## Discussion notes

@@

>>>>

We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs). While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017]. Although enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of the germline genes aberrantly transcribed in _Kdm5c_-KO EpiLCs. One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. However, retinoic acid can only activate _Stra8_ expression when DAZL is present. Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiCs and is both transcribed and translated in _Kdm5c_-KO EpiLCs. Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs. These ectopic germline programs appear to loosely mimic the trajectory of typical germline development, as germline genes important for early germ cell development and meiosis are expressed in _Kdm5c_-KO EpiLC while late spermatogenesis genes are expressed in the mature _Kdm5c_-KO brain. Altogether, this indicates once a subset of germline genes are activated due to loss of direct KDM5C repression in the early embryo, these ectopic germline programs can continue to progress in the background of _Kdm5c_-KO somatic development.

@@

\clearpage
# Notes
## Things to do
- Move dazl to new figure if other staining works
- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
- Add cell culture concentrations to methods
- Motif analysis
	- Discussion - talk about motifs

@@

@@

## Discussion notes

- For other paper:
	- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
	- - Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654

- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
